---
title: "Summary_telomere_mutation"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
```

Read in tables
```{r}
setwd("~/Documents/PhD/Sequencing_results/Mutographs/XX_Signatures/Normal_10/mutation/")
key_table <- read.csv("Chemo_patients.csv", header= T, stringsAsFactors=F)

patients <- key_table$PDID

muts <- read.table(paste0(patients[1],"_mut_dep.txt"), header = T, stringsAsFactors=F)
for (i in 2:length(patients)){
      muts_to_add <- read.table(paste0(patients[i],"_mut_dep.txt"), header = T, stringsAsFactors=F)
      muts <- rbind(muts, muts_to_add)
}

muts$PDID = substr(muts$Sample,1,7)

Summary_All <- left_join(muts, key_table, by = "PDID")

binomial <- read.csv("adapted_vs_binomial.csv", header= T, stringsAsFactors=F)

timepoints <- read.csv("PD47703_timepoints.csv", header= T, stringsAsFactors=F)

```

```{r}
Summary_All$Exposure <- factor(Summary_All$Exposure, levels = c("Normal", "Chemo"))
Summary_All$Bifunctional_alkylating_agent <- factor(Summary_All$Bifunctional_alkylating_agent, levels = c("Cyclophosphamide", "Melphalan and Cyclophosphamide", "Bendamustine","Chlorambucil and Bendamustine","Chlorambucil", "None", "Normal"))
Summary_All$Monofunctional_alkylating_agent <- factor(Summary_All$Monofunctional_alkylating_agent, levels = c("Procarbazine","None","Normal"))
Summary_All$Platinum <- factor(Summary_All$Platinum, levels = c("Oxaliplatin","Carboplatin","Carboplatin and Cisplatin", "None","Normal"))
Summary_All$Antimetabolite <- factor(Summary_All$Antimetabolite, levels = c("5FU","Capecitabine","5FU and Capecitabine","Cytarabine", "None", "Normal"))
Summary_All$Topo_inhibitor <- factor(Summary_All$Topo_inhibitor, levels = c("Daunorubicin","Doxorubicin","Etoposide","Irinotecan", "None","Normal"))
Summary_All$Vinca_alkaloid <- factor(Summary_All$Vinca_alkaloid, levels = c("Vincristine","Vinorelbine and Vinblastine", "None","Normal"))
Summary_All$Cancer_diagnosis <- factor(Summary_All$Cancer_diagnosis, levels = c("AML","Follicular lymphoma","Marginal zone lymphoma", "Lymphoplasmacytic lymphoma", "DLBCL", "Myeloma and Colon cancer", "Hodgkin lymphoma", "Colon cancer", "Lung cancer", "Neuroblastoma", "Normal"))
Summary_All$Years_post_chemo <- factor(Summary_All$Years_post_chemo, levels = c(0,1,2,4,5,6, "None"))
Summary_All$Chemo_class <- factor(Summary_All$Chemo_class, levels = c("Cyclophosphamide","Bendamustine", "Melphalan_cyclophosphamide_oxaliplatin","Chlorambucil","Chlorambucil_procarbazine_bendamustine","Cyclophosphamide_procarbazine","Oxaliplatin","Carboplatin","Carboplatin_cisplatin_cyclophosphamide","None","Normal"))
Summary_All$Years_post_chemo <- factor(Summary_All$Years_post_chemo, levels = c(0,1,2,4,5,6, "None"))

```

Comparison of depth of sequencing between normals and chemotherapy exposed
```{r}
ggplot(Summary_All)+
  theme_bw()+
  scale_colour_manual(values = c(brewer.pal(11,"RdBu")[c(10,8)]))+
  ylim(0,50)+
    theme(text=element_text(size=22))+
    geom_boxplot(data = (Summary_All), aes(group = Exposure, x = Exposure, y =Mean_depth, colour = Exposure))
```

Comparison of depth of sequencing by diagnosis
```{r}
ggplot(Summary_All)+
  theme_bw()+
  theme(text=element_text(size=22))+
    theme(axis.text.x=element_blank())+
  scale_colour_manual(values = c(brewer.pal(12,"Paired")[c(7,8,8,8,5,6,12,4,10,2,3)]))+
  ylim(0,50)+
    geom_boxplot(data = (Summary_All), aes(group = Cancer_diagnosis, x = Cancer_diagnosis, y =Mean_depth, colour = Cancer_diagnosis))
```

Comparison of depth of sequencing by chemotherapy exposure
```{r}
ggplot(Summary_All)+
  theme_bw()+
  theme(text=element_text(size=5))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ylim(0,50)+
    geom_boxplot(data = (Summary_All), aes( x = Chemo_class, y =Mean_depth))
```


Comparison of adapted vs binomial filtering approach
- binomial filtering approach
```{r}
ggplot(binomial)+
  theme_bw()+
  theme(text=element_text(size=22))+
    theme(axis.text.x=element_blank())+
  geom_smooth(data = (binomial), method = "lm",aes(x = Age, y =Number_mutations_binomial), size = 0.5)+
  #scale_colour_manual(values = c(brewer.pal(12,"Paired")[c(7,8,5,6,12,4,10,2)]))+
  ylim(0,2000)+
    geom_boxplot(data = (binomial), aes(group = Age, x = Age, y =Number_mutations_binomial))
```

- adapted filtering approach
```{r}
ggplot(binomial)+
  theme_bw()+
  theme(text=element_text(size=22))+
    theme(axis.text.x=element_blank())+
    geom_smooth(data = (binomial), method = "lm",aes(x = Age, y =Number_mutations), size = 0.5)+
  #scale_colour_manual(values = c(brewer.pal(12,"Paired")[c(7,8,5,6,12,4,10,2)]))+
  ylim(0,2000)+
    geom_boxplot(data = (binomial), aes(group = Age, x = Age, y =Number_mutations))
```


SBS analysis - high burden cases only

```{r}
ggplot(Summary_All)+
  theme_bw()+
  labs(x="Age", y="SBS burden")+
  scale_x_continuous(limits = c(-5,90), breaks = c(0,20,40,60,80))+
  scale_y_continuous(limits = c(0,10000), breaks = c(1000,2000,3000,4000,5000,6000,7000,8000,9000,10000))+
   stat_smooth(data = (Summary_All[Summary_All$Exposure == "Normal",]), method = "lm",aes(x = Age, y =Number_mutations), size = 0.5, fullrange= TRUE)+
  scale_colour_manual(values = c(brewer.pal(12,"Paired")[c(2,6)]))+
    theme(text=element_text(size=18))+
     geom_boxplot(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("AML","Hodgkin lymphoma", "Neuroblastoma", "Normal"),]),aes(group = PDID, x = Age, y =Number_mutations), outlier.alpha = 0)+
     geom_point(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("AML","Hodgkin lymphoma", "Neuroblastoma", "Normal"),]), position = position_dodge(width = 0.75),  aes(group = PDID, x = Age, y =Number_mutations, colour = Exposure))

```


SBS analysis - Low burden cases only

```{r}
ggplot(Summary_All)+
  theme_bw()+
  labs(x="Age", y="SBS burden")+
  scale_x_continuous(limits = c(35,85), breaks = c(40,60,80))+
  scale_y_continuous(limits = c(0,2000), breaks = c(200,400,600,800,1000,1200,1400,1600,1800,2000))+
  stat_smooth(data = (Summary_All[Summary_All$Exposure == "Normal",]), method = "lm",aes(x = Age, y =Number_mutations), size = 0.5, fullrange= TRUE)+
  scale_colour_manual(values = c(brewer.pal(12,"Paired")[c(2,6)]))+
    theme(text=element_text(size=18))+
         geom_boxplot(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("Follicular lymphoma","Marginal zone lymphoma", "Lymphoplasmacytic lymphoma", "DLBCL", "Myeloma and Colon cancer", "Colon cancer", "Lung cancer", "Normal"),]), aes(group = PDID, x = Age, y =Number_mutations), outlier.alpha = 0)+
      geom_point(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("Follicular lymphoma","Marginal zone lymphoma", "Lymphoplasmacytic lymphoma", "DLBCL", "Myeloma and Colon cancer", "Colon cancer", "Lung cancer", "Normal"),]), position = position_dodge(width = 0.75), aes(group = PDID, x = Age, y =Number_mutations, colour = Exposure))

```


Indel analysis - high burden cases only

```{r}
ggplot(Summary_All)+
  theme_bw()+
  labs(x="Age", y="Indel burden")+
  scale_x_continuous(limits = c(-5,90), breaks = c(0,20,40,60,80))+
  scale_y_continuous(limits = c(0,200), breaks = c(0,20,40,60,80,100,120,140,160,180,200))+
  theme(text=element_text(size=18))+
   stat_smooth(data = (Summary_All[Summary_All$Exposure == "Normal",]), method = "lm",aes(x = Age, y =Number_indels), size = 0.5, fullrange= TRUE)+
  scale_colour_manual(values = c(brewer.pal(12,"Paired")[c(2,6)]))+
   geom_boxplot(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("AML","Hodgkin lymphoma", "Neuroblastoma", "Normal"),]),aes(group = PDID, x = Age, y =Number_indels), outlier.alpha = 0)+
     geom_point(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("AML","Hodgkin lymphoma", "Neuroblastoma", "Normal"),]), position = position_dodge(width = 0.75),  aes(group = PDID, x = Age, y =Number_indels, colour = Exposure))
```


Indel analysis - low burden cases only

```{r}
ggplot(Summary_All)+
  theme_bw()+
  labs(x="Age", y="Indel burden")+
  scale_x_continuous(limits = c(35,85), breaks = c(40,60,80))+
  scale_y_continuous(limits = c(0,120), breaks = c(0,20,40,60,80,100,120))+
  theme(text=element_text(size=18))+
   stat_smooth(data = (Summary_All[Summary_All$Exposure == "Normal",]), method = "lm",aes(x = Age, y =Number_indels), size = 0.5, fullrange= TRUE)+
  scale_colour_manual(values = c(brewer.pal(12,"Paired")[c(2,6)]))+
   geom_boxplot(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("Follicular lymphoma","Marginal zone lymphoma", "Lymphoplasmacytic lymphoma", "DLBCL", "Myeloma and Colon cancer", "Colon cancer", "Lung cancer", "Normal"),]), aes(group = PDID, x = Age, y =Number_indels), outlier.alpha = 0)+
      geom_point(data = (Summary_All[Summary_All$Cancer_diagnosis %in% c("Follicular lymphoma","Marginal zone lymphoma", "Lymphoplasmacytic lymphoma", "DLBCL", "Myeloma and Colon cancer", "Colon cancer", "Lung cancer", "Normal"),]), position = position_dodge(width = 0.75), aes(group = PDID, x = Age, y =Number_indels, colour = Exposure))

```

